Single-molecule orientation determination with multiview polarized illumination: modeling and microscope design

### 2017-08-07 paper/paper/paper.tex
Patrick's edits.

\documentclass[10pt]{article}
\usepackage[OE]{express}
\usepackage{float}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\bs}[1]{\boldsymbol{#1}} 
\providecommand{\intinf}{\int_{-\infty}^{\infty}}

\begin{document}
\title{Single molecule orientation determination with multiview polarized
  illumination: modeling and microscope design}

\author{Talon Chandler,\authormark{1,*} Shalin Mehta,\authormark{2} Rudolf
  Oldenbourg,\authormark{2, 3} and Patrick J. La Rivi\`ere\authormark{1}}

\address{\authormark{1}University of Chicago, Department of Radiology, Chicago, IL, USA.\\
  \authormark{2}Marine Biological Laboratory, Bell Center for Regenerative Medicine, Woods Hole, MA, USA.\\
  \authormark{3}Brown University, Department of Physics, Providence, RI, USA.}
\email{*talonchandler@talonchandler.com} %% email address is required

%%%%%%%%%%%%%%%%%%% abstract and OCIS codes %%%%%%%%%%%%%%%%
\begin{abstract*}
  We investigate the use of polarized illumination in multiview microscopes for
  determining the orientation of single molecules. First, we relate the
  orientation of single molecules to measurable intensities in multiview
  microscopes and develop an information theoretic metric---the solid angle
  uncertainty---to compare the ability of multiview microscopes to estimate the
  orientation of single molecules. Next, we compare a broad class of microscopes
  using this metric---single- and dual-view microscopes with varying
  illumination polarization, illumination numerical aperture (NA), detection NA,
  obliquity, asymmetry, and exposure. We find that multiview microscopes are
  better at determining the orientation of single molecules than single-view
  microscopes. We also find that choosing a small illumination NA and a large
  detection NA are good design choices, multiview widefield microscopes can
  benefit from oblique illumination and detection, and asymmetric NA microscopes
  can benefit from exposure asymmetry.
\end{abstract*}

@@

\section{Introduction}
The orientation of single molecules is a valuable reporter for biological
processes. By imaging biological samples with fluorescent probes that are fixed
to known structures, researchers have studied the dynamics of motor proteins
\cite{peterman2001, forkey2003}, DNA \cite{backer2016}, actin \cite{mehta2016},
septin \cite{demay2011, mcquilken2017}, and membranes \cite{anantharam2010}. The
number of techniques available to biologists for measuring the orientation of
single molecules in live cells is continually expanding. We follow a recent
review \cite{backlund2014} and compare three major categories of single molecule
orientation determination techniques:

@@

\textbf{Polarized illumination or detection} techniques exploit the anisotropic
absorption and emission pattern of single molecules to estimate their
orientation. By measuring the total intensity in the detector plane during
several polarization orientations on the illumination or detection path, the
orientation of the molecule can be estimated. These techniques are easy to
implement---changing the illumination or detection polarization is simple with a
polarization splitter \cite{mehta2016} or universal compensator
\cite{oldenbourg1995}---and the reconstruction methods are straightforward
\cite{fourkas2001, mehta2016, backer2016}. The main drawback is speed---these
techniques require the collection of several polarization frames to find the
orientation of a single molecule. Despite the requirement for multiple frames,
polarized illumination is the method of choice for live cell imaging because of
its good combination of speed and ease of implementation. For these reasons,
polarized illumination is the method we consider and extend in this paper.

All single molecule orientation determination methods suffer from some degree of
anisotropic orientation uncertainty, i.e., molecules in some orientations
cannot be oriented as well as molecules in others. In some cases the
orientation cannot be determined at all---the forward model can be degenerate
for specific dipole orientations \cite{fourkas2001, lu2008}. We feel that the
importance of isotropic orientation uncertainty has been underappreciated. To
our knowledge, the only authors who have considered anisotropic orientation
uncertainty have only analyzed a small set of molecule orientations intead of
all orientations \cite{agrawal2012}. Our view is that an ideal technique for
determining the orientation of single molecules can reconstruct the
orientation with a small and nearly uniform uncertainty for all molecule
orientations.

Recently, there has been increasing interest in multiview microscopy techniques
for biological imaging \cite{wu2013, keller2015, wu2016, wu2017}. Multiview
microscopes address two key limitations of single view microscopes. First,
multiview microscopes can achieve nearly isotropic resolution compared to the
poor axial resolution of single view microscopes. Second, multiview microscopes
can easily incorporate selective illumination to reduce phototoxicity. For
example, compare light-sheet illumination to wide-field illumination where
most of the illumination light excites out of focus molecules. These
advantages make multiview microscopes good candidates for imaging live
biological specimens for long periods.

@@

\subsection{Absorption Efficiency}\label{excitation}
Consider polarized widefield K\"ohler illumination incident on a single
molecule at the focal point with the optical axis along the $\mh{z}$ axis. In
this section we will calculate the \emph{absorption efficiency} in this
geometry---the fraction of the incident power that is absorbed by the molecule.

We start with the unit electric field transmitted by the polarizer,
\begin{align}
 \hat{\mb{E}}_{\text{exc}} = \cos\phi_{\text{exc}}\hat{\mb{x}} +
\sin\phi_{\text{exc}}\hat{\mb{y}},\label{eq:excfield}
\end{align}
where $\phi_{\text{exc}}$ is the angle between the transmission axis of the
excitation polarizer and the $\mh{x}$ axis. Next, we model the action of the an
ideal, aplanatic, polarization preserving illumination objective by multiplying
the electric field vector by a direction dependent rotation matrix,
  \begin{align}
  \mb{R}(\mh{r}) &= \begin{bmatrix} \cos\theta\cos^2\phi + \sin^2\phi & (\cos\theta -1)\sin\phi\cos\phi & -\sin\theta\cos\phi\\ (\cos\theta - 1)\sin\phi\cos\phi & \cos\theta\sin^2\phi + \cos^2\phi & -\sin\theta\sin\phi \\ \sin\theta\cos\phi& \sin\theta\sin\phi & \cos\theta \end{bmatrix}
  \end{align}
  where
  \begin{align}
   \hat{\mb{r}} = \sin\theta\cos\phi\hat{\mb{x}} + \sin\theta\sin\phi\hat{\mb{y}} + \cos\theta\hat{\mb{z}} 
  \end{align}
  is a unit vector that points from the origin to the point where the incident
  electric field intersects the Gaussian sphere of the illumination
  objective. Next, we find the power that excites the molecule by taking the
  dot product of the rotated incident field $\mb{R}(\mh{r})\mh{E}_{\text{exc}}$
  and the absorption dipole moment
\begin{align}
  \hat{\bs{\mu}}_{\text{abs}} = \sin\Theta\cos\Phi\hat{\mb{x}} + \sin\Theta\sin\Phi\hat{\mb{y}} + \cos\Theta\hat{\mb{z}}
\end{align}
then taking its modulus squared. If the incident fields are spatially incoherent
then we can find the total power that excites the molecule by integrating over the
illuminated region of the Gaussian sphere $\Omega$ where
\begin{align}
  \Omega &= \{\phi, \theta\ |\ 0 < \phi \leq 2\pi, 0 < \theta \leq \alpha\}, \\
  \alpha &= \arcsin\left(\frac{\text{NA}}{n}\right), \label{eq:alpha}
\end{align}
NA is the illumination numerical aperture, and $n$ is the index of refraction of
the medium.

Finally, we find the fraction of the incident power absorbed by the molecule
by dividing the total absorbed power by the total incident power. The complete
expression for the absorption efficiency in vector notation is
\begin{align}
  \eta_{\text{abs}} = \frac{\int_{\Omega}d\mh{r}|\hat{\bs{\mu}}_{\text{abs}}\cdot\mb{R}(\mh{r})\mh{E}_{\text{exc}}|^2}{\int_{\Omega}d\mh{r}}\label{eq:abs}. 
\end{align}

We substitute equations \ref{eq:excfield}--\ref{eq:alpha} into equation
\ref{eq:abs}, evaluate the integrals, and simplify to express the absorption
efficiency in scalar notation as
\begin{align}
  \eta_{\text{abs}} &= D\{A + B\sin^{2}{\Theta} + C\sin^{2}{\Theta} \cos{[2 (\Phi - \phi_{\text{exc}}})]\}\label{eq:scalarabs}
\end{align}
where
\begin{subequations}
\begin{align}
  A &= \frac{1}{4} - \frac{3}{8} \cos{\alpha } + \frac{1}{8} \cos^{3}{\alpha }\\
  B &= \frac{3}{16} \cos{\alpha } - \frac{3}{16} \cos^{3}{\alpha }\\
  C &= \frac{7}{32} - \frac{3}{32} \cos{\alpha } - \frac{3}{32} \cos^{2}{\alpha } - \frac{1}{32} \cos^{3}{\alpha}\\
  D &= \frac{4}{3(1 - \cos\alpha)}.
\end{align}\label{eq:coefficients}%
\end{subequations}

At the beginning of this section we assumed that the molecule was located at
the focal point of the condenser. If the illumination objective is aplanatic
then we can relax this assumption and the above expressions are valid for
molecules anywhere in the focal plane.

We also assumed that we used incoherent K\"ohler illumination. If we illuminate
the back aperture with a weakly focused laser beam and scan the laser beam
slowly---if the scan velocity is much less than the beam waist divided by the
coherence time---we can ignore longitudinal excitation and the molecule will
be excited as if a single plane was incident. Therefore, we can find the
absorption efficiency of a molecule under weakly focused scanned laser
illumination by taking the limit of equation \ref{eq:scalarabs} as
$\alpha \rightarrow 0$ giving
\begin{align}
  \eta_{\text{abs}} &= \sin^2\Theta\cos^2(\Phi - \phi_{\text{exc}})
\end{align}
which is recognizable as Malus' law generalized to three dimensions. Notice that
the absorption efficiency can take its a maximum range of values
$\eta_{\text{abs}} \in [0, 1]$ when $\alpha=0$.

@@

Also note that the detection efficiency can take its a maximum range of values
when $B$ takes its maximum value, i.e. when
$\alpha=\arccos\left(\frac{1}{\sqrt{3}}\right) \approx 54.7^{\circ}$. When $n = 1.33$,
this condition corresponds to $\text{NA} \approx 1.1$.

\subsection{Oblique Illumination and Detection}\label{oblique}
In sections \ref{excitation} and \ref{detection} we assumed that both the
illumination and detection objectives had $\mh{z}$ aligned optical axes. To extend
the forward model to oblique optical axes, we express the dipole orientation in
rotated coordinates using the following expressions
\begin{align}
    \Theta' &= \arccos\left(\sin\psi\cos\Phi\sin\Theta + \cos\psi\cos\Theta\right)\label{eq:thetap}\\
  \Phi' &=
          \begin{cases}
            \arccos\left(\dfrac{\cos\psi\cos\Phi\sin\Theta - \sin\psi\cos\Theta}{\sqrt{1 - (\sin\psi\cos\Phi\sin\Theta + \cos\psi\cos\Theta)^2}}\right) \ \ \ &0 \leq \Phi < \pi  \\
            -\arccos\left(\dfrac{\cos\psi\cos\Phi\sin\Theta - \sin\psi\cos\Theta}{\sqrt{1 - (\sin\psi\cos\Phi\sin\Theta + \cos\psi\cos\Theta)^2}}\right) \ \ \ &-\pi \leq \Phi < 0\\
          \end{cases}
\end{align}
where $\psi$ is the angle of a right handed rotation about the $\mh{y}$ that maps the $\mh{z}$ axis onto the new optical axis.

@@

Note our distinction between \emph{frames} and \emph{views}. Frames refer to
single intensity measurements with fixed illumination, detection, and
polarization geometries. Views refers to the detection objectives, and multiview
microscopes use multiple detection objectives.

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{single-frame}
\caption{Representative examples of single-frame microscopes.\newline \newline
  \textbf{Columns left to right:} 1) schematics of single-frame microscopes
  where the solid line encloses the illumination solid angle, the dashed line
  encloses the detection solid angle, and the arrow indicates the transmission
  axis of the illumination polarizer; 2) the absorption efficiency, see equation
  \ref{eq:scalarabs}; 3) the detection efficiency, see equation
  \ref{eq:scalardet}; (4) the total efficiency, the product of the absorption and
  detection efficiencies.\newline \newline \textbf{Rows top to bottom:} 1)
  epi-illumination ($\text{NA} = 1.1$) with $x$-polarized light and
  epi-detection (NA = 1.1); 2) orthogonal illumination ($\text{NA} = 0.8$) and
  detection ($\text{NA} = 0.8$); 3) oblique illumination ($\text{NA} = 0.8$) and
  detection ($\text{NA} = 0.8$). All simulations use $n=1.33$.}
  \label{fig:single-frame}
\end{figure}

@@

We also consider several dual-view designs. We evaluate two-view designs that
allow for illumination and detection from both objectives (see \cite{wu2013} for
a dual-view design). In all cases we consider the effect of varying the
illumination and detection numerical aperture. We also consider asymmetric NA
designs (see \cite{wu2017} for an asymmetric NA design) and the effect of
sample exposure.

@@

For each microscope design and molecule orientation we calculate the Fisher
information matrix. If the intensity is Poisson distributed the
Fisher information matrix is given by
\begin{align}
  \mb{F} &= \sum_{k=1}^N \frac{1}{I_k}
  \begin{bmatrix}
    \dfrac{\partial I_k}{\partial \Theta}\dfrac{\partial I_k}{\partial \Theta}&\dfrac{\partial I_k}{\partial \Theta}\dfrac{\partial I_k}{\partial \Phi}\\\\
    \dfrac{\partial I_k}{\partial \Theta}\dfrac{\partial I_k}{\partial \Phi}&\dfrac{\partial I_k}{\partial \Phi}\dfrac{\partial I_k}{\partial \Phi}\\    
  \end{bmatrix}
\end{align}
where $I_k$ is the intensity measured in the $k$th frame of an $N$ frame
microscope.

A common way to evaluate the ability to estimate the parameters $\Theta$ and
$\Phi$ from the data is to calculate the Cramer-Rao lower bound (CRLB) for each
parameter \cite{kay1993}. The CRLBs are given by the diagonal elements of the
inverse of the Fisher information matrix, and they give the minimum variance of
an unbiased estimator for each parameter. Agrawal et. al. used the square root
of the product of the CRLBs multiplied by the Jacobian determinant,
$\sin\Theta\sqrt{\mb{F}^{-1}_{0,0}\mb{F}^{-1}_{1,1}}$, to find the area of
uncertainty in parameter space \cite{agrawal2012}.

CRLBs and the associated area of uncertainty are parametrization dependent,
i.e., if we choose a different coordinate system these metrics will change. We
would like to compare microscope designs without choosing a parametrization, so
instead we use
\begin{align}
  \sigma_{\Omega} &\equiv \sin\Theta\sqrt{\text{det}\{\mb{F}^{-1}\}}
\end{align}
as our evaluation metric. We take the determinant of the inverse Fisher
information matrix---a parametrization independent value called the
\emph{generalized variance} \cite{anderson1958}---take the square root, then
multiply by the Jacobian determinant, $\sin\Theta$. We call $\sigma_{\Omega}$
the \emph{solid-angle uncertainty} because it has units of steradians and is a
parametrization independent measure of the orientation uncertainty.

For each multi-frame microscope we calculate $\sigma_{\Omega}$ at 10,000
approximately equally spaced points on the unit sphere. A desirable microscope
design will have a small solid-angle uncertainty that is uniform for all
molecule orientations. The most straightforward way to find the location and
scale of the solid-angle uncertainty is to use the mean and variance,
respectively. Because the solid-angle uncertainty admits extremely large values
that can change the mean and variance dramatically depending on the sample
points, we use the median and median absolute deviation (MAD)---the median of
the data's absolute difference from the median---as robust replacements for the
mean and variance.

@@

We have considered single- and dual-view microscopes in a variety of cases, but
more multiview microscopes can be analyzed with the framework we have
developed. Wu et. al. have developed a three view design with a third objection
below the cover slip\cite{wu2016}. Light-field detection schemes are also a type
of multiview microscope\cite{levoy2006}, and we plan to extend our analysis to
this case in future work.

>>>>

\documentclass[10pt]{article}
\usepackage[OE]{express}
\usepackage{float}
\usepackage{booktabs}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\bs}[1]{\boldsymbol{#1}} 
\providecommand{\intinf}{\int_{-\infty}^{\infty}}

\begin{document}
\title{Single-molecule orientation determination with multiview polarized
  illumination: modeling and microscope design}

\author{Talon Chandler,\authormark{1,*} Shalin Mehta,\authormark{2} Hari Shroff,\authormark{3,4} Rudolf Oldenbourg,\authormark{2, 5} and Patrick J. La Rivi\`ere\authormark{1, 4}}
\address{\authormark{1}University of Chicago, Department of Radiology, Chicago, Illinois 60637, USA\\
  \authormark{2}Marine Biological Laboratory, Bell Center for Regenerative Medicine, Woods Hole, Massachusetts, USA\\
  \authormark{3}Section on High Resolution Optical Imaging, National Institute
  of Biomedical Imaging and Bioengineering, National Institutes of Health,
  Bethesda, Maryland 20892, USA\\
  \authormark{4}Whitman Center, Marine Biological Laboratory, Woods Hole,
  Massachusetts 02543, USA\\
  \authormark{5}Brown University, Department of Physics, Providence, Rhode
  Island 02912, USA}
\email{*talonchandler@talonchandler.com} %% email address is required

%%%%%%%%%%%%%%%%%%% abstract and OCIS codes %%%%%%%%%%%%%%%%
\begin{abstract*}
  We investigate the use of polarized illumination in multiview microscopes for
  determining the orientation of single molecules. First, we relate the
  orientation of single molecules to measurable intensities in multiview
  microscopes and develop an information theoretic metric---the solid angle
  uncertainty---to compare the ability of multiview microscopes to estimate the
  orientation of single molecules. Next, we compare a broad class of microscopes
  using this metric---single- and dual-view microscopes with varying
  illumination polarization, illumination numerical aperture (NA), detection NA,
  obliquity, asymmetry, and exposure. We find that multiview microscopes are
  better at determining the orientation of single molecules than single-view
  microscopes. We also find that choosing a small illumination NA and a large
  detection NA are good design choices, that multiview widefield microscopes can
  benefit from oblique illumination and detection, and that asymmetric NA
  microscopes can benefit from exposure asymmetry.
\end{abstract*}

@@

\section{Introduction}
The orientation of single molecules is a valuable reporter for biological
processes. By imaging biological samples with fluorescent probes that are fixed
to known structures, researchers have studied the dynamics of motor proteins
\cite{peterman2001, forkey2003}, DNA \cite{backer2016}, actin \cite{mehta2016},
septin \cite{demay2011, mcquilken2017}, and membranes \cite{anantharam2010}. The
number of techniques available to biologists for measuring the orientation of
single molecules in live cells is continually expanding. A recent review
\cite{backlund2014} identified three major categories of single molecule
orientation determination techniques:

@@

\textbf{Polarized illumination or detection} techniques exploit the anisotropic
absorption and emission pattern of single molecules to estimate their
orientation. By measuring the total intensity in the detector plane while
varying polarizer orientations on the illumination or detection path, the
orientation of the molecule can be estimated. These techniques are easy to
implement---changing the illumination or detection polarization is simple with a
polarization splitter \cite{mehta2016} or universal compensator
\cite{oldenbourg1995}---and the reconstruction methods are straightforward
\cite{fourkas2001, mehta2016, backer2016}. The main drawback is that these
techniques require several polarization frames to estimate the orientation of a
single molecule. Despite the requirement for multiple frames, polarized
illumination or detection methods are good choices for live cell imaging because
of their good combination of speed and ease of implementation.

The choice remains---is it best to put polarizers on the illumination path or
the detection path? Placing a polarizer in the illumination path lengthens
acquisition time because the polarizer needs to be changed for each polarization
frame, but polarized illumination provides the largest intensity differences
between molecules in different orientations. Placing polarizers in the detection
path doesn't slow down acquisition (multiple polarization frames can be
collected at once with a polarization splitter), but polarized detection
provides a smaller difference in intensity between molecules in different
orientations than polarized illumination. We are most interested in finding the
orientation of molecules with high precision, so polarized illumination is the
method we consider and extend in this paper.

All single molecule orientation determination methods suffer from some degree of
anisotropic orientation uncertainty, i.e., some molecular orientations cannot be
determined as precisely as others. In some cases the orientation cannot be
determined at all---the forward model can be degenerate for specific dipole
orientations \cite{fourkas2001, lu2008}. We feel that the importance of
isotropic orientation uncertainty has been underappreciated. To our knowledge,
the only authors who have considered anisotropic orientation uncertainty have
only analyzed a small set of molecular orientations instead of all orientations
\cite{agrawal2012}. Our view is that an ideal technique for determining the
orientation of single molecules can reconstruct the orientation with a small and
nearly uniform uncertainty for all molecule orientations.

Recently, there has been increasing interest in multiview microscopy techniques
for biological imaging \cite{wu2013, keller2015, wu2016, wu2017}. Multiview
microscopes offer two major advantages over single-view microscopes. First,
multiview microscopes can achieve nearly isotropic resolution compared to the
poor axial resolution of single-view microscopes. Second, multiview microscopes
can use light-sheet illumination to reduce phototoxicity without requiring
dedicated illumination optics. For example, if two orthogonal objectives are
focused on the same point, then both objectives can alternate roles as the
light-sheet illumination path and as the detection path. Light-sheet
illumination offers a major reduction of phototoxicity because a light sheet
only illuminates in focus regions while single-view microscopes illuminate out
of focus regions. Together, these advantages make multiview microscopes good
candidates for imaging live biological specimens for long periods.

@@

Note our distinction between a \emph{frame} and a \emph{view}. A frame refers to
a single intensity measurement with a fixed polarizer setting, illumination
path, and detection path. Acquiring additional frames entails varying the
polarizer setting, illumination path, or detection path. A view refers to a
detection path, and multiview microscopes use multiple detection paths.

We use roman type for scalars, e.g., $\phi$, $\theta$; bold lowercase type for
vectors, e.g., $\mb{r}, \bs{\mu}$; hats for unit vectors, e.g.,
$\mh{r}, \hat{\bs{\mu}}$; and bold capital type for matrices, e.g.,
$\mb{R}, \mb{F}$.

\subsection{Absorption Efficiency}\label{excitation}
In this section we will calculate the \emph{absorption efficiency} of a single
molecule---the fraction of the incident power that is absorbed by the
molecule. Our approach is inspired by Fourkas \cite{fourkas2001}, but here we
calculate the absorption efficiency instead of the detection efficiency.

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{frames}
\caption{Coordinate systems for a) the absorption dipole moment
  $\hat{\bs{\mu}}_{\text{abs}}$, b) the polarizer transmission axis
  $\hat{\mb{e}}_{\text{exc}}$, and c) the dummy integration vector
  $\hat{\mb{r}}$.}
\label{fig:coordinates}
\end{figure}

We consider a single molecule with an absorption dipole moment
$\hat{\bs{\mu}}_{\text{abs}}$ that we express in spherical coordinates $\Theta$
and $\Phi$ (see Figure \ref{fig:coordinates}a)) as
\begin{align}
  \hat{\bs{\mu}}_{\text{abs}}(\Theta, \Phi) = \sin\Theta\cos\Phi\hat{\mb{x}} + \sin\Theta\sin\Phi\hat{\mb{y}} + \cos\Theta\hat{\mb{z}} \label{eq:mu}.
\end{align}
We place the molecule at the focal point of an ideal, aplanatic,
polarization-preserving objective with the optical axis aligned with the
$\mh{z}$ axis. Finally, we place a polarizer behind the objective with a
variable transmission axis $\hat{\mb{e}}_{\text{exc}}$ that we express as
\begin{align}
 \hat{\mb{e}}_{\text{exc}}(\phi_{\text{exc}}) = \cos\phi_{\text{exc}}\hat{\mb{x}} +
\sin\phi_{\text{exc}}\hat{\mb{y}},\label{eq:excfield}
\end{align}
where $\phi_{\text{exc}}$ is the angle between the transmission axis of the
polarizer and the positive $\mh{x}$ axis (see Figure
\ref{fig:coordinates}b)). We illuminate the polarizer with a beam of collimated
light propagating along the optical axis, so the electric field incident on the
objective is parallel with the transmission axis of the polarizer.

We denote the illuminated region of the objective by $\Omega$. If we use
unit vectors $\hat{\mb{r}}$ expressed in spherical coordinates $\theta$ and
$\phi$ (see Figure \ref{fig:coordinates}c)) as
  \begin{align}
   \hat{\mb{r}}(\theta, \phi) = \sin\theta\cos\phi\hat{\mb{x}} + \sin\theta\sin\phi\hat{\mb{y}} + \cos\theta\hat{\mb{z}}, 
  \end{align}
  then the region $\Omega$ can be expressed as
  \begin{align}
  \Omega = \{\phi, \theta\ |\ 0 < \phi \leq 2\pi, 0 < \theta \leq \alpha\},
  \end{align}
  where $\alpha$ is the angle between the optical axis and the most oblique
  illuminating ray. Equivalently, $\alpha$ can be expressed in terms of NA using
  \begin{align}
    \text{NA} = n\sin\alpha, 
  \end{align}
  where $n$ is the index of refraction of the sample medium. Note that the NA of
  the objective is always greater than or equal to the illumination NA because
  we can under fill the back aperture of the objective.

  The objective focuses the collimated illumination by applying a position-dependent
  rotation to the electric field which we model by multiplying the incident electric
  field $\hat{\mb{e}}$ by a position-dependent rotation matrix
  \begin{align}
  \mb{R}(\mh{r}) &= \begin{bmatrix} \cos\theta\cos^2\phi + \sin^2\phi & (\cos\theta -1)\sin\phi\cos\phi & -\sin\theta\cos\phi\\ (\cos\theta - 1)\sin\phi\cos\phi & \cos\theta\sin^2\phi + \cos^2\phi & -\sin\theta\sin\phi \\ \sin\theta\cos\phi& \sin\theta\sin\phi & \cos\theta \end{bmatrix} \label{eq:rot_mat}.
  \end{align}

  To find the absorption efficiency $\eta_{\text{abs}}$ of a single molecule if
  it were illuminated by a single ray, we take the unit incident electric field
  $\hat{\mb{e}}$, take the dot product with the absorption dipole moment, then
  take the modulus squared
    \begin{align}
      \eta_{\text{abs}} &= |\hat{\bs{\mu}}_{\text{abs}}\cdot\mh{e}_{\text{exc}}|^2. \label{eq:singleray}
    \end{align}

    To find the absorption efficiency of a single molecule under incoherent
    focused light, we integrate over all rays and divide by the total incident
    power which gives the vector expression
\begin{align}
  \eta_{\text{abs}} = \frac{\int_{\Omega}d\mh{r}|\hat{\bs{\mu}}_{\text{abs}}\cdot\mb{R}(\mh{r})\mh{e}_{\text{exc}}|^2}{\int_{\Omega}d\mh{r}}\label{eq:abs}. 
\end{align}

We substitute equations \ref{eq:mu}--\ref{eq:rot_mat} into equation
\ref{eq:abs}, evaluate the integrals, and simplify to express the absorption
efficiency in scalar notation as
\begin{align}
  \eta_{\text{abs}} &= D\{A + B\sin^{2}{\Theta} + C\sin^{2}{\Theta} \cos{[2 (\Phi - \phi_{\text{exc}}})]\}\label{eq:scalarabs}
\end{align}
where
\begin{subequations}
\begin{align}
  A &= \frac{1}{4} - \frac{3}{8} \cos{\alpha } + \frac{1}{8} \cos^{3}{\alpha }\\
  B &= \frac{3}{16} \cos{\alpha } - \frac{3}{16} \cos^{3}{\alpha }\\
  C &= \frac{7}{32} - \frac{3}{32} \cos{\alpha } - \frac{3}{32} \cos^{2}{\alpha } - \frac{1}{32} \cos^{3}{\alpha}\\
  D &= \frac{4}{3(1 - \cos\alpha)}.
\end{align}\label{eq:coefficients}%
\end{subequations}

At the beginning of this section we assumed that the molecule was located at the
focal point of the condenser. If the illumination objective is aplanatic and we
use K\"ohler illumination, then we can relax this assumption and the above
expressions are valid for molecules anywhere in the focal plane.

We also assumed that we used incoherent illumination, but we can extend these
results to light-sheet illumination that uses a scanned coherent laser. If we
illuminate the back aperture with a weakly focused laser beam and scan the laser
beam slowly---if the scan velocity is much less than the beam waist radius
divided by the coherence time---we can ignore longitudinal excitation and the
molecule will be excited as if a single plane wave was incident. Therefore, we
can find the absorption efficiency of a molecule under weakly focused scanned
laser illumination by taking the limit of equation \ref{eq:scalarabs} as
$\alpha \rightarrow 0$ (or by plugging equation \ref{eq:mu} and
\ref{eq:excfield} into equation \ref{eq:singleray}) giving
\begin{align}
  \eta_{\text{abs}} &= \sin^2\Theta\cos^2(\Phi - \phi_{\text{exc}}),
\end{align}
which is recognizable as Malus' law generalized to three dimensions. This means
that light-sheet illumination is approximately equivalent to widefield
illumination with a low NA. Notice that the absorption efficiency can take its
maximum range of values ($\eta_{\text{abs}} \in [0, 1]$) when $\alpha=0$.

@@

Also note that the detection efficiency can take its maximum range of values
when $B$ takes its maximum value, i.e. when
$\alpha=\arccos\left(\frac{1}{\sqrt{3}}\right) \approx 54.7^{\circ}$.

\subsection{Oblique Illumination and Detection}\label{oblique}
In sections \ref{excitation} and \ref{detection} we assumed that both the
illumination and detection objectives had $\mh{z}$ aligned optical axes, i.e.,
the same objective was used for illumination and detection. To extend the
forward model to oblique optical axes which allows for non-coincident
illumination and detection objectives, we express the dipole orientation in
rotated coordinates using the following expressions
\begin{align}
    \Theta' &= \arccos\left(\sin\psi\cos\Phi\sin\Theta + \cos\psi\cos\Theta\right)\label{eq:thetap}\\
  \Phi' &=
          \begin{cases}
            \arccos\left(\dfrac{\cos\psi\cos\Phi\sin\Theta - \sin\psi\cos\Theta}{\sqrt{1 - (\sin\psi\cos\Phi\sin\Theta + \cos\psi\cos\Theta)^2}}\right) \ \ \ &0 \leq \Phi < \pi  \\
            -\arccos\left(\dfrac{\cos\psi\cos\Phi\sin\Theta - \sin\psi\cos\Theta}{\sqrt{1 - (\sin\psi\cos\Phi\sin\Theta + \cos\psi\cos\Theta)^2}}\right) \ \ \ &-\pi \leq \Phi < 0\\
          \end{cases}
\end{align}
where $\psi$ is the angle of a right handed rotation about the $\mh{y}$ axis
that maps the $\mh{z}$ axis onto the new optical axis.

@@

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{single-frame}
\caption{Representative examples of single-frame microscopes.\newline \newline
  \textbf{Columns left to right:} 1) schematics of single-frame microscopes
  where the solid line encloses the illumination solid angle, the dashed line
  encloses the detection solid angle, and the arrow indicates the transmission
  axis of the illumination polarizer; 2) the absorption efficiency, see equation
  \ref{eq:scalarabs}; 3) the detection efficiency, see equation
  \ref{eq:scalardet}; (4) the total efficiency, the product of the absorption
  and detection efficiencies.\newline \newline \textbf{Rows top to bottom:} 1)
  coincident illumination ($\text{NA} = 1.1$ with $x$-polarized light) and
  detection (NA = 1.1); 2) non-coincident orthogonal illumination
  ($\text{NA} = 0.8$) and detection ($\text{NA} = 0.8$); 3) non-coincident
  135${}^{\circ}$-separated illumination ($\text{NA} = 0.8$) and detection
  ($\text{NA} = 0.8$). All simulations use $n=1.33$.}
  \label{fig:single-frame}
\end{figure}

@@

We also consider several dual-view designs. We evaluate two-view designs that
allow for illumination and detection from both objectives (see \cite{wu2013} for
a dual-view design). In all cases we consider the effect of varying the
illumination and detection numerical aperture. We also consider asymmetric NA
designs (see \cite{wu2017} for an asymmetric NA design) and the effect of
asymmetric sample exposures. Note that we use the word \emph{symmetry} to refer
to dual-view designs with objectives that are identical and the word
\emph{asymmetry} to refer to dual-view designs with objectives that have a
different NA or sample exposure.

@@

A common way to evaluate the ability to estimate the parameters $\Theta$ and
$\Phi$ from the data is to calculate the Cramer-Rao lower bound (CRLB) for each
parameter \cite{kay1993}. The CRLBs are given by the diagonal elements of the
inverse of the Fisher information matrix, and they give the minimum variance of
an unbiased estimator for each parameter. For each microscope design and
molecule orientation we calculate the Fisher information matrix. If the
intensity is Poisson distributed the Fisher information matrix is given by
\begin{align}
  \mb{F} &= \sum_{k=1}^N \frac{1}{I_k}
  \begin{bmatrix}
    \dfrac{\partial I_k}{\partial \Theta}\dfrac{\partial I_k}{\partial \Theta}&\dfrac{\partial I_k}{\partial \Theta}\dfrac{\partial I_k}{\partial \Phi}\\\\
    \dfrac{\partial I_k}{\partial \Theta}\dfrac{\partial I_k}{\partial \Phi}&\dfrac{\partial I_k}{\partial \Phi}\dfrac{\partial I_k}{\partial \Phi}\\    
  \end{bmatrix}
\end{align}
where $I_k$ is the intensity measured in the $k$th frame of an $N$-frame
microscope. Agrawal et. al. used the square root of the product of the CRLBs
multiplied by the Jacobian determinant,
$\sin\Theta\sqrt{\mb{F}^{-1}_{0,0}\mb{F}^{-1}_{1,1}}$, to find the area of
uncertainty in parameter space \cite{agrawal2012}.

CRLBs and the associated area of uncertainty are parametrization dependent,
i.e., if we choose a different coordinate system these metrics will change. We
would like to compare microscope designs without choosing a parametrization, so
instead we use
\begin{align}
  \sigma_{\Omega} &\equiv \sin\Theta\sqrt{\text{det}\{\mb{F}^{-1}\}}
\end{align}
as our evaluation metric. We calculate the determinant of the inverse Fisher
information matrix---a parametrization-independent value called the
\emph{generalized variance} \cite{anderson1958}---take the square root, then
multiply by the Jacobian determinant, $\sin\Theta$. We call $\sigma_{\Omega}$
the \emph{solid-angle uncertainty} because it has units of steradians and is a
parametrization independent measure of the orientation uncertainty.

For each multi-frame microscope we calculate $\sigma_{\Omega}$ at 10,000
approximately equally spaced points on the unit sphere. A desirable microscope
design will have a small solid-angle uncertainty that is uniform for all
molecule orientations. The most straightforward way to find the location and
scale of the solid-angle uncertainty is to use the mean and variance,
respectively. However, because the solid-angle uncertainty admits extremely
large values that can change the mean and variance dramatically depending on the
sample points, we use the median and median absolute deviation (MAD)---the
median of the data's absolute difference from the median---as robust
alternatives to the mean and variance.

@@

We have considered single- and dual-view microscopes in a variety of cases, but
more multiview microscopes can be analyzed with the framework we have
developed. Wu et. al. have developed a three-view design with a third objective
below the cover slip\cite{wu2016}. Light-field detection schemes are also a type
of multiview microscope\cite{levoy2006}, and we plan to extend our analysis to
this case in future work.

Although we have only considered single-molecule orientation determination in
this work, polarized illumination techniques can be applied to find the
distribution of ensembles of molecules \cite{mehta2016, backer2016}. In future
work we will investigate the ability of multiview polarized illumination
microscopes to estimate the orientation and distribution of ensembles of
molecules.
